---
title: "2024_04_03_drug_antibody_targets"
format: html
editor: visual
---

Trying some new methods to discover targets for drug and antibody treatment among the upregulated genes shared by both HOS and OS17 exposed to tumor supernatant.

# Libraries

```{r}
suppressPackageStartupMessages({
library(tidyverse)
library(readr)
})
```

# Set up directories

```{r}
result.dir <- "2024_04_03_results/"

if(!exists(result.dir)){
  dir.create(result.dir)
}
```

# Read in data

```{r}
drug.df <- read_delim("data/repurposing.drug.data.txt", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE) |> 
  drop_na(Target)

ab.df <- read_csv("data/fda.approved.antibodies.csv")

gene.df <- read_csv("2024_03_08_results/HOS_OS17_shared_DE_genes.csv")
```

Starting with the drug dataframe, because the gene target names seem consistent with our results. The antibody Target column may need additional formatting.

First, some drugs have multiple targets. I need to split those up into separate columns.

```{r}
drug.df |> 
  separate_longer_delim(Target, delim = ",")
```





























